Interactive Data Exploration with Smart Drill-Down (Extended Version)

### 2016-12-11 TableSummarization/journalVersion/smartDrillDown.tex
hector comments

\noindent
\textbf{Runtime analysis:}
\mrj{Our algorithm calls the function for finding best marginal rule $k$ times. Thus the total computational cost is about $k$ times the cost of computing the best marginal rule. Let $c_n$ be the number of counts in $C_n$ at the end of the $n^{th}$ iteration (that is, after eliminating rules that have no chance of having the highest marginal value). Also, let $c_0 = 1$.}

\mrj{The cost of the finding the best marginal rule can be split into three parts: (i) Enumerating all new rules to add to $C_n$ at the start of the $n^{th}$ iteration. (ii) Determining which of the rules in $C_n$ to eliminate (iii) Finding counts and marginal values for the remaining $c_n$ rules in $C_n$. The number of super-rules we can consider for (i) can be no larger than $c_{n-1}|T|$, since each rule in $C_{n-1}$ can give us at most $|T|$ new candidate super-rules for $C_n$. The cost of (ii) is $O(|C_n|)$. Finally, the cost of (iii) is again $c_n|T|$, since we have two nested for loops over $T$ and $C_n$ respectively. Thus, our total cost is $O(|T|\sum_i c_i)$.}

\mrj{In the worst case, we may need to evaluate all possible rules; since the total possible number of rules is $|T||C|$ where $C$ is the set of columns, our cost is at most $O(k|C||T|^2))$. However, note that this worst case is unlikely to occur when we use the Size or Bits weighting function, because the count of a rule decreases rapidly with its size, making it unlikely that we will need to find counts for its super-rules.}

\mrj{In fact, if we assume that values in columns occur independently of each other, and that the most frequent value occurs $x < 1$ fraction of the time, then the number of candidate rules is bounded by $c_i \leq c_1 x^i$. This gives us $\sum_i c_i \leq c_1\sum_i x^i \leq c_1/(1-x)$.
Thus our cost is now only $O(k|T|c_1/(1-x))$ for some constant $x$. If the number of distinct values in the table ($c_1$), $k$ and $x$ are taken to be constants (i.e. they do not scale with table size), then our cost is $O(|T|)$.}

\noindent
\textbf{Approximation ratio:}
\mrj{If $m_w$ is higher than the actual weight of every rule in the best rule list, then the only approximation ratio we incur is that in the greedy approximation, which is a multiplicative factor of up to $1 - 1/e$. On the other hand, if our $m_w$ estimate is smaller than the actual maximum weight $m_w^*$ of a rule in the best list, then we may end up picking a sub-optimal rule in the best marginal rule function.}

\mrj{We now bound the sub-optimality of our chosen rule. Suppose the best rule to pick is $r^*$, and we pick rule $r$. Since we estimated $r^*$'s weight to be at most $m_w$, we would have estimated $r^*$ to have marginal value $\text{Est}(r^*) = \sum_{t \in r^*}(m_w - w(t))$ where $t \in r^*$ iterates over all tuples covered by $r^*$, and $w(t)$ is the weight of the current highest weight rule covering $t$. Since we picked $r$ over $r^*$, it means $\text{MarginalValue}(r) \geq \text{Est}(r^*)$. Suppose $m$ is the highest weight of a rule selected in the previous steps of BRS (so $w(t) \leq m$ for all $t$). Then, the actual marginal value of $r^*$ is $\leq \frac{m_w - m}{m_w^* - m}\text{Est}(r^*)$. Thus, our pick of $r$ is at most a $\frac{m_w - m}{m_w^* - m}$ factor worse than the optimal pick. If we always choose  our $m_w$ such that we have $\frac{m_w - m}{m_w^* - m} \geq w$, for some constant $w < 1$, then our final score is guaranteed to be within a $w(1-1/e)$ factor of the optimal score.}

@@

We describe our technique to efficiently allocate memory to different samples, so as to maximize the probability that we can respond to the next user operation without accessing the hard disk in Section~\ref{sec:sampling_algorithms} through Section~\ref{sec:alternative-convex}. 
Then, in Section~\ref{sec:sample-handler}, we describe a component of our system, called the {\em SampleHandler}, which is responsible for creating and maintaining samples of the table in memory, subject to user specified memory constraints. The SampleHandler maintains multiple samples corresponding to different parts of the table, which can be used depending on which rule the user decides to expand next. Finally, we mention some additional optimizations we can make, and
describe how we can set the minimum sample size required from the SampleHandler.

@@

\papertext{
\subsection{Alternative Convex-Optimization based solution}\label{sec:alternative-convex}
We noted earlier that Problem~\ref{prob:sample-sizes} is NP-Hard, but can be approximately solved with an additional simplifying assumption regarding the $S(r_1, r_2)$ values. Instead of making this simplification, we can make the problem convex (and hence tractable) with two different simplifications. The first simplification is, we modify our objective function to use hinge-loss instead of a step function. That is, our new objective function to maximise is $$\sum_{r^{\prime} \in L} p_{r^{\prime}}\textrm{min}\left(1, \frac{ess(r^{\prime})}{minSS}\right)$$
Here we assume that it is acceptable to run our algorithm on samples smaller than $minSS$, though we still prefer bigger sample sizes upto $minSS$. The other simplification we make is assuming that sample sizes are real numbers instead of integers. After determining optimal sample sizes, we can round them up to get integer sample sizes. This will increase the memory usage by at most $|U|$, the number of nodes in displayed tree, which is negligible compared to the memory capacity $M$, or $minSS$.

@@

\mrj{We now analyze a parametric family of weighting functions that generalizes our functions from Section~\ref{sec:weighting} and provides intuition for them. Let $W(r) = \left(\sum_{c \in C} o_{r,c}w_c\right)^k$, where the $w_c$'s and $k$ are parameters of the weighting function, and $o_{r,c}$ ($o$ stands for occurrence) is $1$ if $r(c) \neq \star$ and $0$ otherwise. The Size and Bits weighting functions can be seen as special cases of the above function with $k=1$ and with $w_c$ equal to $1$ for Size and $\text{log}(|c|)$ for Bits.}

\mrj{We want to estimate the Score contribution of a rule according to this weight function, which requires estimating the count of the rule. Suppose that for each column $c$, the most frequent value occurs a $f_c$ fraction of the time. If the probability of a value occurring in a column of a tuple is independent of other values in the tuple, the $Count$ of rule $r$ is approximately $|T|\Pi_{c \in C}f_c^{o_{r,c}}$. Thus the score of a rule set containing $r$ alone can be approximated by:
$$S(r) = \left(\sum_{c \in C} o_{r,c}w_c\right)^k|T|\Pi_{c \in C}f_c^{o_{r,c}}$$
}

\mrj{Now consider the rule that has the highest weight times count according to the above weighting function. We can approximate this rule by relaxing $o_{r,c}$ to a real number; that is by maximizing the estimated $S(r)$ subject to $0 \leq o_{r,c} \leq 1$. According to the KKT conditions~\cite{Boyd:2004:CO:993483}, at the optimum $Weight \times Count$, the partial derivative of $W(r)$ with respect to $o_{r,c}$ must be $\leq 0$ for each $c$ where $o_{r,c} = 0$, equal to $0$ if $0 < o_{r,c} < 1$ and $\geq 0$ otherwise. For each $c$, we have:
$$\frac{\partial S(r)}{\partial o_{r,c}} = S(r)(\frac{kw_c}{\sum_{c \in C} o_{r,c}w_c} + \ln f_c)$$}

\mrj{Thus, if $x = \frac{k}{\sum_{c \in C} o_{r,c}w_c}$, then the maximum scoring rule should instantiate columns where $\frac{\ln f_c}{w_c} > x$ and some of the ones with $\frac{\ln f_c}{w_c} = x$. Thus, for Size weighting, the highest score rule should select columns with high values of $\ln f_c$, i.e., the highest frequency values. In contrast, Bits weighting will select columns with highest values of $\log_{|c|}(f_c)$, i.e., values that are frequent relative to the number of distinct values in the column.}

\mrj{If we want the score to treat all columns equally, then the derivative must be $0$ for all $c$. This implies that for all $c$, we must have $w_c \propto \ln f_c$. Notice that Bits satisfies the above criterion if we assume that values in each column are uniformly distributed (that is, $f_c = 1/|c|$). This provides an intuitive justification for why we might want to use the Bits weighting function.}

\mrj{We can also estimate the size of the optimal rule under the above weighting function. Specifically, the weighted fraction of columns with non-$\star$ values is given by $\frac{\sum_{c\in C} o_{r,c}w_c}{\sum_{c \in C} w_c}$ which equals $-k/(\sum_{c \in C} \ln f_c)$. Thus, if we wanted our highest score rule to have $s$ fraction of the columns get instantiated, we can achieve that by setting $k = -s\sum_{c \in C} \ln f_c$.}

\mrj{The above reasoning also lets us estimate the $m_w$ parameter we need to use. Specifically, the highest score rule above has weight $(\frac{-\ln f_c}{kw_c})^k$.}

\mrj{We can use the above estimate to guess the size of the max scoring rule for the Bits weighting function. But note that in practice, values are not distributed uniformly (so $f_c$ is actually higher than $1/|c|$) and are not independent of each other, due to correlations. As a result, our estimate of the maximum size is a significant under-estimate. In our experiments (Section~\ref{sec:exp_qualitative}), we observe that use the Bits weighting function actually gives us rules of size $3$ on our datasets, while our estimate above is approximately $0.74$. As expected, the values displayed by the rules are ones that occur much more frequently relative to other values in that column.}
}
%This evaluates to $1.44$, while the observed value in our experiments is $10$.

\subsection{Dealing with Numerical Attributes}\label{sec:extensions-numerical}
\mrj{Our framework assumes that all attributes are categorical. Attributes that have a large domain tend to have fewer tuples per value, and hence don't appear in rule summaries. Thus our algorithm does not summarize information about numerical attributes. But it can easily be modified to do so by bucketizing a numerical attribute and treating the bucket id as a categorical attribute. This is already done in our MD dataset, where numerical attributes like age are divided into buckets ($18-24$, $25-34$ and so on).}

\vspace{-5pt}
\subsection{Using Sum instead of Count}\label{sec:extensions-sum}
\mrj{Throughout the paper, we define the total score of a rule-list using the marginal counts of rules in the list, and display the count of each rule in our table summary. However, if we have a numerical column (i.e. a `measure' column) in the table, it is straightforward to extend our summary to the `Sum' aggregate over that column instead. We can define $Sum$ and $MSum$ for a rule analogously to $Count$ and $MCount$, and $Score$ can be modified to use $MSum$ instead of $MCount$. We can then modify Algorithm~\ref{algo:best-rule-set} to maximize the new score by replacing $Count(r)$ by $Sum(r)$ and computing sum and marginal sum instead of count and marginal count in each pass over the table.}
}

>>>>

\noindent
\textbf{Runtime analysis:}
Our algorithm calls the function for finding best marginal rule $k$ times. Thus the total computational cost is about $k$ times the cost of computing the best marginal rule. Let $c_n$ be the number of counts in $C_n$ at the end of the $n^{th}$ iteration (that is, after eliminating rules that have no chance of having the highest marginal value). Also, let $c_0 = 1$.

The cost of the finding the best marginal rule can be split into three parts: (i) Enumerating all new rules to add to $C_n$ at the start of the $n^{th}$ iteration. (ii) Determining which of the rules in $C_n$ to eliminate (iii) Finding counts and marginal values for the remaining $c_n$ rules in $C_n$. The number of super-rules we can consider for (i) can be no larger than $c_{n-1}|T|$, since each rule in $C_{n-1}$ can give us at most $|T|$ new candidate super-rules for $C_n$. The cost of (ii) is $O(|C_n|)$. Finally, the cost of (iii) is again $c_n|T|$, since we have two nested for loops over $T$ and $C_n$ respectively. Thus, our total cost is $O(|T|\sum_i c_i)$.

In the worst case, we may need to evaluate all possible rules; since the total possible number of rules is $|T||C|$ where $C$ is the set of columns, our cost is at most $O(k|C||T|^2))$. However, note that this worst case is unlikely to occur when we use the Size or Bits weighting function, because the count of a rule decreases rapidly with its size, making it unlikely that we will need to find counts for its super-rules.

If fact, say that values in columns occur independently, and let $x$ be the frequency of the most common value (note $x<1$). Then the number of candidate rules is bounded by $c_i \le c_1 x^i$, where $c_1$ is the number of distinct values in the table. This gives us $\sum_i c_i \leq c_1\sum_i x^i \leq c_1/(1-x)$.
Thus our cost is now only $O(k|T|c_1/(1-x))$ for some constant $x$. If $c_1$, $k$ and $x$ are taken to be constants (i.e. they do not scale with table size), then our cost is $O(|T|)$.

\noindent
\textbf{Approximation ratio:}
If $m_w$ is higher than the actual weight of every rule in the best rule list, then the only approximation ratio we incur is that in the greedy approximation, which is a multiplicative factor of up to $1 - 1/e$. On the other hand, if our $m_w$ estimate is smaller than the actual maximum weight $m_w^*$ of a rule in the best list, then we may end up picking a sub-optimal rule in the best marginal rule function.

We now bound the sub-optimality of our chosen rule, for the case where $m_w < m_w^*$. Suppose the best rule to pick is $r^*$, and we pick rule $r$. Since we estimated $r^*$'s weight to be at most $m_w$, we would have estimated $r^*$ to have marginal value $\text{Est}(r^*) = \sum_{t \in r^*}(m_w - W(t))$ where $t \in r^*$ iterates over all tuples covered by $r^*$, and $w(t)$ is the weight of the current highest weight rule covering $t$. Since we picked $r$ over $r^*$, it means $\text{MarginalValue}(r) \geq \text{Est}(r^*)$. Suppose $m$ is the highest weight of a rule selected in the previous steps of BRS (so $w(t) \leq m$ for all $t$). Then, the actual marginal value of $r^*$ is $\leq \frac{m_w - m}{m_w^* - m}\text{Est}(r^*)$. Thus, our pick of $r$ is at most a $\frac{m_w - m}{m_w^* - m}$ factor worse than the optimal pick. If we always choose  our $m_w$ such that we have $\frac{m_w - m}{m_w^* - m} \geq a$, for some constant $a < 1$, then our final score is guaranteed to be within a $a(1-1/e)$ factor of the optimal score.

@@

We describe our technique to efficiently allocate memory to different samples, so as to maximize the probability that we can respond to the next user operation without accessing the hard disk in Section~\ref{sec:sampling_algorithms}\techreporttext{ through Section~\ref{sec:alternative-convex}}. 
Then, in Section~\ref{sec:sample-handler}, we describe a component of our system, called the {\em SampleHandler}, which is responsible for creating and maintaining samples of the table in memory, subject to user specified memory constraints. The SampleHandler maintains multiple samples corresponding to different parts of the table, which can be used depending on which rule the user decides to expand next. Finally, we mention some additional optimizations we can make, and
describe how we can set the minimum sample size required from the SampleHandler.

@@

\techreporttext{
\subsection{Alternative Convex-Optimization based solution}\label{sec:alternative-convex}
We noted earlier that Problem~\ref{prob:sample-sizes} is NP-Hard, but can be approximately solved with an additional simplifying assumption regarding the $S(r_1, r_2)$ values. Instead of making this simplification, we can make the problem convex (and hence tractable) with two different simplifications. The first simplification is, we modify our objective function to use hinge-loss instead of a step function. That is, our new objective function to maximise is $$\sum_{r^{\prime} \in L} p_{r^{\prime}}\textrm{min}\left(1, \frac{ess(r^{\prime})}{minSS}\right)$$
Here we assume that it is acceptable to run our algorithm on samples smaller than $minSS$, though we still prefer bigger sample sizes upto $minSS$. The other simplification we make is assuming that sample sizes are real numbers instead of integers. After determining optimal sample sizes, we can round them up to get integer sample sizes. This will increase the memory usage by at most $|U|$, the number of nodes in displayed tree, which is negligible compared to the memory capacity $M$, or $minSS$.

@@

We now analyze a parametric family of weighting functions that generalizes our functions from Section~\ref{sec:weighting} and provides intuition for them. Let $W(r) = \left(\sum_{c \in C} o_{r,c}w_c\right)^k$, where the $w_c$'s and $k$ are parameters of the weighting function, and $o_{r,c}$ ($o$ stands for occurrence) is $1$ if $r(c) \neq \star$ and $0$ otherwise. The Size and Bits weighting functions can be seen as special cases of the above function with $k=1$ and with $w_c$ equal to $1$ for Size and $\text{log}(|c|)$ for Bits.

We want to estimate the Score contribution of a rule according to this weight function, which requires estimating the count of the rule. Suppose that for each column $c$, the most frequent value occurs a $f_c$ fraction of the time. If the probability of a value occurring in a column of a tuple is independent of other values in the tuple, the $Count$ of rule $r$ is approximately $|T|\Pi_{c \in C}f_c^{o_{r,c}}$. Thus the score of a rule set containing $r$ alone can be approximated by:
$$S(r) = \left(\sum_{c \in C} o_{r,c}w_c\right)^k|T|\Pi_{c \in C}f_c^{o_{r,c}}$$

Now consider the rule that has the highest weight times count according to the above weighting function. We can approximate this rule by relaxing $o_{r,c}$ to a real number; that is by maximizing the estimated $S(r)$ subject to $0 \leq o_{r,c} \leq 1$. According to the KKT conditions~\cite{Boyd:2004:CO:993483}, at the optimum $Weight \times Count$, the partial derivative of $W(r)$ with respect to $o_{r,c}$ must be $\leq 0$ for each $c$ where $o_{r,c} = 0$, equal to $0$ if $0 < o_{r,c} < 1$ and $\geq 0$ otherwise. For each $c$, we have:
$$\frac{\partial S(r)}{\partial o_{r,c}} = S(r)(\frac{kw_c}{\sum_{c \in C} o_{r,c}w_c} + \ln f_c)$$

Thus, if $x = \frac{k}{\sum_{c \in C} o_{r,c}w_c}$, then the maximum scoring rule should instantiate columns where $\frac{\ln f_c}{w_c} > x$ and some of the ones with $\frac{\ln f_c}{w_c} = x$. Thus, for Size weighting, the highest score rule should select columns with high values of $\ln f_c$, i.e., the highest frequency values. In contrast, Bits weighting will select columns with highest values of $\log_{|c|}(f_c)$, i.e., values that are frequent relative to the number of distinct values in the column.

If we want the score to treat all columns equally, then the derivative must be $0$ for all $c$. This implies that for all $c$, we must have $w_c \propto \ln f_c$. Notice that Bits satisfies the above criterion if we assume that values in each column are uniformly distributed (that is, $f_c = 1/|c|$). This provides an intuitive justification for why we might want to use the Bits weighting function.

We can also estimate the size of the optimal rule under the above weighting function. Specifically, the weighted fraction of columns with non-$\star$ values is given by $\frac{\sum_{c\in C} o_{r,c}w_c}{\sum_{c \in C} w_c}$ which equals $-k/(\sum_{c \in C} \ln f_c)$. Thus, if we wanted our highest score rule to have $s$ fraction of the columns get instantiated, we can achieve that by setting $k = -s\sum_{c \in C} \ln f_c$.

The above reasoning also lets us estimate the $m_w$ parameter we need to use. Specifically, the highest score rule above has weight $(\frac{-\ln f_c}{kw_c})^k$.

We can use the above estimate to guess the size of the max scoring rule for the Bits weighting function. But note that in practice, values are not distributed uniformly (so $f_c$ is actually higher than $1/|c|$) and are not independent of each other, due to correlations. As a result, our estimate of the maximum size is a significant under-estimate. In our experiments (Section~\ref{sec:exp_qualitative}), we observe that use the Bits weighting function actually gives us rules of size $3$ on our datasets, while our estimate above is approximately $0.74$. As expected, the values displayed by the rules are ones that occur much more frequently relative to other values in that column.
}
%This evaluates to $1.44$, while the observed value in our experiments is $10$.

\subsection{Dealing with Numerical Attributes}\label{sec:extensions-numerical}
Our framework assumes that all attributes are categorical. Attributes that have a large domain tend to have fewer tuples per value, and hence don't appear in rule summaries. Thus our algorithm does not summarize information about numerical attributes. But it can easily be modified to do so by bucketizing a numerical attribute and treating the bucket id as a categorical attribute. This is already done in our MD dataset, where numerical attributes like age are divided into buckets ($18-24$, $25-34$ and so on).

\vspace{-5pt}
\subsection{Using Sum instead of Count}\label{sec:extensions-sum}
Throughout the paper, we define the total score of a rule-list using the marginal counts of rules in the list, and display the count of each rule in our table summary. However, if we have a numerical column (i.e. a `measure' column) in the table, it is straightforward to extend our summary to the `Sum' aggregate over that column instead. We can define $Sum$ and $MSum$ for a rule analogously to $Count$ and $MCount$, and $Score$ can be modified to use $MSum$ instead of $MCount$. We can then modify Algorithm~\ref{algo:best-rule-set} to maximize the new score by replacing $Count(r)$ by $Sum(r)$ and computing sum and marginal sum instead of count and marginal count in each pass over the table.}
